A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-21 Manuscript/PNAS-Adams-02.MLC.Rmd
mlc comments

The appeal of Pagel's $\lambda$ and Blomberg's $\kappa$ as descriptive statistics is that they provide a basis for interpreting "weak" versus "strong" phylogenetic signal; i.e., small versus large values of $\hat{\lambda}$ or $\kappa$, respectively, in a comparative sense [@DeMeester2019; @Pintanel2019;@Su2019]. Nonetheless, an important question that has yet to be considered is whether such comparisons are analytically appropriate, and whether these statistics are, or can be, converted to effect sizes for comparative analyses across datasets. To be statistics representing phylogenetic signal, they should have reliable distributional properties, which could be revealed with simulation experiments. For instance, as a proportional random variable bounded by $0$ and $1$, we might expect that $\hat{\lambda}$ follows a Bernoulli distribution (**add ref**); i.e., branch lengths in a tree are scaled proportionally to the probability that data arise from a BM process. Given a known $\lambda$ value used to generate random data on a tree, we would also expect  that the mean of an empirical sampling distribution of $\hat{\lambda}$ would approximately equal $\lambda$; the dispersion of $\hat{\lambda}$ would be largest at intermediate values of $\lambda$, $\hat{\lambda}$ would be predictable over the range of $\lambda$ with respect to tree size; the distribution of $\hat{\lambda}$ would be symmetric at intermediate values of $\lambda$ and more skewed toward values of 0 or 1; and that the distribution of $\hat{\lambda}$ will be more platykurtic at intermediate values of $\lambda$, becoming more leptokurtic toward 0 and 1 (**add same ref**). Prior work [@Munkemuller_et_al2012] seems to support some of these conjectures, based superficially on statistical moments for a given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of ref. [@Munkemuller_et_al2012]). However, because the "strength of Brownian motion" was simulated as a varied weighted-average of data simulated on trees with $\lambda=0$ and $\lambda=1$ and not as prescribed values of $\lambda$ [@Munkemuller_et_al2012], interpretation of these patterns is challenging.

By contrast, for Blomberg's $\kappa$, which is positively unbounded, we might expect that for any $\lambda$ used to generate data, estimates of $\kappa$ might follow a normal distribution, with values distributed symmetrically about the input value. This attribute seemed less reasonable based on the simulations performed by Münkemüller et al. [-@Munkemuller_et_al2012], which suggested that distributions were positively skewed and that Blomberg's $\kappa$ might not behave as a statistic that follows a normal distribution. However, because their simulations used a weighted combination of simulated phylogenetic signal strengths, strong inferences are not possible (and distributional attributes were not the intended result of their simulations). Thus, for both Pagel's $\lambda$ or Blomberg's $\kappa$, evaluation of statistical moments across a range of $\lambda$ used to generate data would be valuable for adjudicating the reliability of these statistics as effect sizes.  Furthermore, the expected values of these statistics appear to vary with tree size [@Munkemuller_et_al2012], making comparisons across studies challenging. Therefore, transformation of these statistics into $Z$-scores would allow evaluation of the efficacy of each statistic to yield effect sizes that could be used for comparisons of the strength of phylogenetic signal across traits and lineages.

@@

**Lambda ($\lambda$) estimates of phylogenetic signal are innacurate.** Computer simulations reveal that for $\hat{\lambda}$, the distributional expectations of a Bernoulli variable were mostly upheld. First, the mean value of $\hat{\lambda}$ increases as $\lambda$ increases, but it is negatively-biased (particularly for small tree sizes), and is consistently less than the input $\lambda$ value across most of its range (Fig. 1 black line). Second, the standard deviation of $\hat{\lambda}$ is largest at intermediate values of $\lambda$ and smallest at extreme values (Fig.1 red line), implying that the precision in estimating $\lambda$ varies across the range of input values. Additionally, standard deviations of $\hat{\lambda}$ are negatively associated with tree size, and for tress of 128 species or less, $\hat{\lambda}$ are quite variable, except for cases when $\lambda$ is near or equal to $1$. Third, the distributions of $\hat{\lambda}$ are not normal across its range, but become increasingly skewed at more extreme values of $\lambda$ (Fig. 1 blue line). For small tree sizes, it is also clear that distributions are more platykurtic at intermediate values of $\hat{\lambda}$. Taken together these results reveal that $\hat\lambda$ inconsistently estimates  phylogenetic signal, both across tree sizes and across the range of input values. Additional simulations (Supplemental Information) reveal that incorporating $\hat\lambda$ in PGLS anova and regression does not adversely affect the statistical properties of PGLS parameter estimation or model evaluation (type I error, power, bias in coefficients). Thus, it is reasonable to incorporate $\hat\lambda$ in PGLS as a parameter for tuning the degree of phylogenetic signal in the dependent variables during the analysis. However, the statistical properties shown in Fig. 1 demonstrate that $\lambda$ is unsuitable as an effect size for measuring the strength of phylogenetic signal in data, and thus $\lambda$ should not be used for comparing phylogenetic signal across datasets.

@@

where $\kappa_{obs}$ is the observed phylogenetic signal, and $\hat\mu_\kappa$ and $\hat\sigma_\kappa$ are the mean and standard deviation of the empirical sampling distribution of $\kappa$ obtained via permutation. The empirical sampling distribution of $\kappa$ is first transformed via Box-Cox to better adhere to the assumption of normality.

@@

To demonstrate the utility of $\hat{Z}_12$, we compared  $Z_{\kappa}$ for two ecologically-relevant traits in plethodontid salamander (Fig. 4): surface area to volume ratios (SA:V) and relative body width ($\frac{BW}{SVL}$) [@Baken2019; @Baken2020]. While both traits contained significant phylogenetic signal, tests based on $\hat{Z}_{12}$ revealed that the degree of phylogenetic signal was significantly stronger in SA:V ($\hat{Z}_{12}=4.13$; $P=0.000036$: Fig. 4). Biologically, this observation may be interpreted by the fact that tropical species -- which form a monophyletic group within plethodontids -- display greater variation in SA:V, which covaries with disparity in their climatic niches [@Baken2020]. Thus, greater phylogenetic signal in SA:V is to be expected.

>>>>

The appeal of Pagel's $\lambda$ and Blomberg's $\kappa$ as descriptive statistics is that they provide a basis for interpreting "weak" versus "strong" phylogenetic signal; i.e., small versus large values of $\hat{\lambda}$ or $\kappa$, respectively, in a comparative sense [@DeMeester2019; @Pintanel2019;@Su2019]. Nonetheless, an important question that has yet to be considered is whether such comparisons are analytically appropriate, and whether these statistics are, or can be, converted to effect sizes for comparative analyses across datasets. To be statistics representing phylogenetic signal, they should have reliable distributional properties, which could be revealed with simulation experiments. For instance, as a proportional random variable bounded by $0$ and $1$, we might expect that $\hat{\lambda}$ is a random variable that follows the distribution of a Bernoulli probability parameter [@Forbes2011]; i.e., branch lengths in a tree are scaled proportionally to the probability that data arise from a BM process. Given a known $\lambda$ value used to generate random data on a tree, we would also expect  that the mean of an empirical sampling distribution of $\hat{\lambda}$ would approximately equal $\lambda$; the dispersion of $\hat{\lambda}$ would be largest at intermediate values of $\lambda$, $\hat{\lambda}$ would be predictable over the range of $\lambda$ with respect to tree size; the distribution of $\hat{\lambda}$ would be symmetric at intermediate values of $\lambda$ and more skewed toward values of 0 or 1; and that the distribution of $\hat{\lambda}$ will be more platykurtic at intermediate values of $\lambda$, becoming more leptokurtic toward 0 and 1 [@Forbes2011]. Prior work [@Munkemuller_et_al2012] seems to support some of these conjectures, based superficially on statistical moments for a given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of ref. [@Munkemuller_et_al2012]). However, because the "strength of Brownian motion" was simulated as a varied weighted-average of data simulated on trees with $\lambda=0$ and $\lambda=1$ and not as prescribed values of $\lambda$ [@Munkemuller_et_al2012], interpretation of these patterns is challenging.

By contrast, for Blomberg's $\kappa$, which is positively unbounded, we might expect that for any $\lambda$ used to generate data, estimates of $\kappa$ might be a random variable that follows a normal distribution, with values distributed symmetrically about the input value [@Forbes2011]. This attribute seemed less reasonable based on the simulations performed by Münkemüller et al. [-@Munkemuller_et_al2012], which suggested that distributions were positively skewed and that Blomberg's $\kappa$ might not behave as a statistic that follows a normal distribution. However, because their simulations used a weighted combination of simulated phylogenetic signal strengths, strong inferences are not possible (and distributional attributes were not the intended result of their simulations). Thus, for both Pagel's $\lambda$ or Blomberg's $\kappa$, evaluation of statistical moments across a range of $\lambda$ used to generate data would be valuable for adjudicating the reliability of these statistics as effect sizes.  Furthermore, the expected values of these statistics appear to vary with tree size [@Munkemuller_et_al2012], making comparisons across studies challenging. Therefore, transformation of these statistics into $Z$-scores would allow evaluation of the efficacy of each statistic to yield effect sizes that could be used for comparisons of the strength of phylogenetic signal across traits and lineages.

@@

**Lambda ($\lambda$) estimates of phylogenetic signal are innacurate.** Computer simulations reveal that for $\hat{\lambda}$, the distributional expectations of a Bernoulli variable were mostly upheld. First, the mean value of $\hat{\lambda}$ increases as $\lambda$ increases, but it is negatively-biased (particularly for small tree sizes), and is consistently less than the input $\lambda$ value across most of its range (Fig. 1 black line). Second, the standard deviation of $\hat{\lambda}$ is largest at intermediate values of $\lambda$ and smallest at extreme values (Fig.1 red line), implying that the precision in estimating $\lambda$ varies across the range of input values. Additionally, standard deviations of $\hat{\lambda}$ are negatively associated with tree size, and for tress of 128 species or less, $\hat{\lambda}$ are quite variable, except for cases when $\lambda$ is near or equal to $1$. Third, the distributions of $\hat{\lambda}$ are not normal across its range, but become increasingly skewed at more extreme values of $\lambda$ (Fig. 1 blue line). For small tree sizes, it is also clear that distributions are more platykurtic at intermediate values of $\hat{\lambda}$. Taken together these results reveal that $\hat\lambda$ inconsistently estimates  phylogenetic signal, both across tree sizes and across the range of input values. Additional simulations (Supplemental Information) reveal that incorporating $\hat\lambda$ in PGLS ANOVA and regression does not adversely affect the statistical properties of PGLS parameter estimation or model evaluation (type I error, power, bias in coefficients). Thus, it is reasonable to incorporate $\hat\lambda$ in PGLS as a parameter for tuning the degree of phylogenetic signal in the dependent variables during the analysis. However, the statistical properties shown in Fig. 1 demonstrate that $\lambda$ is unsuitable as an effect size for measuring the strength of phylogenetic signal in data, and thus $\lambda$ should not be used for comparing phylogenetic signal across datasets.

@@

where $\kappa_{obs}$ is the observed phylogenetic signal, and $\hat\mu_\kappa$ and $\hat\sigma_\kappa$ are the mean and standard deviation of the empirical sampling distribution of $\kappa$ obtained via permutation. The empirical sampling distribution of $\kappa$ is first transformed via a Box-Cox transformation to better adhere to the assumption of normality.

@@

To demonstrate the utility of $\hat{Z}_{12}$, we compared  $Z_{\kappa}$ for two ecologically-relevant traits in plethodontid salamander (Fig. 4): surface area to volume ratios (SA:V) and relative body width ($\frac{BW}{SVL}$) [@Baken2019; @Baken2020]. While both traits contained significant phylogenetic signal, tests based on $\hat{Z}_{12}$ revealed that the degree of phylogenetic signal was significantly stronger in SA:V ($\hat{Z}_{12}=4.13$; $P=0.000036$: Fig. 4). Biologically, this observation may be interpreted by the fact that tropical species -- which form a monophyletic group within plethodontids -- display greater variation in SA:V, which covaries with disparity in their climatic niches [@Baken2020]. Thus, greater phylogenetic signal in SA:V is to be expected.
